---
title: "Preterm - Birth"
subtitle: "Logistic Regression"
author: "Jomel Meeko Manzano"
date: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true 
---    
    

```{r global_options, include=FALSE}
# these are some optional settings that will change how some features look
# you do not need to change them.
knitr::opts_chunk$set(out.width = "50%", out.height="50%", fig.align="center", warning=FALSE, message=FALSE)
```


```{r, include=FALSE}


hoslem=function (obs, pred, g = 0)
  {
  if (g == 0)
  g = round(min(length(obs)/10, 10))
  ord <- order(pred)
  obs.o <- obs[ord]
  pred.o <- pred[ord]
  interval = cut(pred.o, quantile(pred.o, 0:g/g), include.lowest = TRUE)
  counts = xtabs(formula = cbind(obs.o, pred.o) ~ interval)
  centers <- aggregate(formula = pred.o ~ interval, FUN = "mean")
  pear.res <- (counts[, "obs.o"] - counts[, "pred.o"])/sqrt(counts[,
  "pred.o"])
  pearson <- sum(pear.res^2)
  if (any(counts[, "pred.o"] < 5))
  warning("Some expected counts are less than 5. Use smaller number of groups")
  p = 1 - pchisq(pearson, g - 2)
  results <- as.data.frame(cbind(counts[, "obs.o"], counts[,
  "pred.o"], centers$pred.o, pear.res))
  colnames(results) <- c("obs.o", "pred.o", "avg mean", "pearson resid")
  cat("Hosmer-Lemeshow test with", g, "bins", "\n", "Pearson Stat = ",
  pearson, "\n", "p = ", p, "\n \n")
  return(results)
  }

```


```{r, echo=FALSE}

library(tidyverse)
library(medicaldata)
library(knitr)
library(kableExtra)
library(car)
library(logistf)
library(ResourceSelection)
library(mice)
# --- Notes --- 

# Remember this is a Clinical Trial we do not need to backward selection

# --- Referrences ---
# Soneji S, Beltrán-Sánchez H. Association of Maternal Cigarette Smoking and Smoking Cessation With Preterm Birth. JAMA Netw Open. 2019;2(4):e192514. doi:10.1001/jamanetworkopen.2019.2514



```
In this phase of the analysis, we concentrate on two secondary outcomes: preterm birth, defined as delivery before 37 weeks, and low birthweight, with the threshold set at below 2500g. It's essential to note that these outcomes are secondary in the context of the clinical trial, where the power calculation was based on gestational age at delivery. This serves as a reminder that our statistical power might be limited for these secondary outcomes, potentially impacting our ability to detect statistical significance.

We are specifically interested in what preterm  birth and low birthweight is associated with. 

### Data Cleaning 
```{r}

opt_raw = opt

opt_logistic <- opt_raw %>%
  # turn space to NA in ended_37_wk
  mutate(
    cat_birthweight = factor(ifelse(Birthweight < 2500, 0, 1), 
                             levels = c(0, 1)),
    across(where(is.factor), ~na_if(as.character(.), "   "))
  ) %>%
  select(
    # Outcome
    Preg.ended...37.wk, cat_birthweight,
    # Center
    Clinic,
    # demographic
    Age, White, Black, Hisp, Education, Prev.preg, 
    # Coexisting medical condition
    Diabetes, Hypertension, Drug.Add, Use.Alc, Use.Tob, # eating disorder
    # dental status ?
    ## missing number of natural teeth
    N.qualifying.teeth, BL..BOP, BL.PD.avg
  )  %>%
  mutate_if(is.character, as.factor)
```

### Missing Value Imputation
```{r}
colSums(is.na(opt_logistic))/nrow(opt_logistic)

opt_logistic = opt_logistic %>%
  # delete missing outcome
  filter(!is.na(Preg.ended...37.wk) & !is.na(cat_birthweight))

imputation = mice(
  opt_logistic,
  m = 10,
  maxit = 10,
  seed = 1,
  method = "pmm"
)
opt_logistic = complete(imputation,1)
```


### Logistic Regression

#### Part 1 - Preterm or Not Preterm

##### Model Fitting 
```{r, echo=FALSE}


# --- Fit Base Model ---
  
model.preterm <- glm(as.factor(Preg.ended...37.wk) ~ . - cat_birthweight, 
                     data = opt_logistic, 
          family = binomial, na.action = na.exclude)

model.preterm.summary <- summary(model.preterm)
model.preterm.summary$coefficients %>% 
  round(digits = 3) %>% kable()

# --- Variable Selection --- 

model.preterm.reduced <- glm(as.factor(Preg.ended...37.wk) ~ . - 
                               cat_birthweight - Drug.Add - Use.Alc  - Use.Tob, 
                             data = opt_logistic, 
          family = binomial, na.action = na.exclude)

model.preterm.reduced.summary <- summary(model.preterm.reduced)
model.preterm.reduced.summary$coefficients %>% 
  round(digits = 3) %>% kable()

```
We have two models above. The bigger model has these variables **Clinic**, **Age**, **Ethnicity**, **Education**, **Number of previous pregnancies**, **Diabetes**, **Hypertension**, **Alcohol use**, **Number of qualifying teeth**, **Percentage of sites bleeding on probing at baseline**, **Whole-mouth average pocket depth at baseline (mm)**, **Self-reported participant history of drug addiction** and **Self-reported participants history of alcohol use**. The reduced model does not have drug addiction and alcohol use variables. We are just curious if these vices are contributing to these secondary outcomes in the sample. 



##### Likelihood Ratio Test
```{r, echo=FALSE}
anova(model.preterm.reduced, model.preterm, 
      test = "LRT")

```
- $H_{0}$: There is no statistical difference between the reduced and full model.
- $H_{A}$: There is statistical difference between the reduced and full model. 
- P-value: 0.878
- We fail to reject the null hypothesis at significance level 0.05.
- The full model does not perform better and we decided to go with the simpler model that does not have alcohol, drug and tobacco use. 



```{r, echo=FALSE}
compareCoefs(model.preterm,model.preterm.reduced) %>% kable()
```


We also examined the coefficients of both models to determine if there were any significant changes. As depicted in the table above, the coefficients remain stable and fall within the standard errors. Additionally, there are no coefficients that have changed signs.



##### Final Model
```{r, echo=FALSE}


# --- Goodness of Fit ---


opt_logistic$Preg.ended...37.wk <- as.factor(opt_logistic$Preg.ended...37.wk )

generalhoslem::logitgof(opt_logistic$Preg.ended...37.wk,
                       model.preterm.reduced$fitted.values)

# student residuals

std.res <- rstandard(model.preterm.reduced) # student residual
std.res[std.res == max(std.res)] 

# Outlier Test
outlierTest(model.preterm.reduced)

```
- We are using a Hosmer-Lemeshow for Goodness-of-Fit
- $H_{0}$: There is no significant difference between the observed and expected values of the binary response variable across different groups or categories defined by the predictor variables.
- $H_{A}$: There is a significant difference between the observed and expected values of the binary response variable across different groups or categories defined by the predictor variables.
- We fail to reject the null hypothesis at a significance level of 0.05. This suggests that the model fits well.
- We also conducted an outlier test to identify any data points that may be influencing our model. We observed that participant 121 had a high studentized residual; however, this outlier was only present in one out of 809 participants and is unlikely to significantly impact our analysis.

##### Conclusion 
- Interpretations:
  - Intercept is not interpretable because age will never be zero and qualifying teeth will almost never be zero. 
  - Individuals with **hypertension (HypertensionY = Yes)** are estimated to have approximately 4.28 times higher odds of experiencing preterm birth compared to those without hypertension, controlling for other variables in the model.
  - Individuals with **diabetes (DiabetesYes  = Yes)** are estimated to have approximately 2.82 times higher odds of experiencing preterm birth compared to those without hypertension, controlling for other variables in the model.



#### Part 2 - Birthweight 

```{r, echo=FALSE}


#cat_birthweight -> Reference level is 0 meaning low birthweight. 

# --- Full Model ---

opt_logistic$cat_birthweight <- as.factor(opt_logistic$cat_birthweight)

model.birthweight <- glm(cat_birthweight ~ . - Preg.ended...37.wk , 
                         data = opt_logistic, family = binomial, 
                         na.action = na.exclude)

model.birthweight.summary <- summary(model.birthweight)
model.birthweight.summary$coefficients %>% 
  round(digits = 3) %>% kable()

# --- Reduced Model ---

model.birthweight.reduced <- glm(cat_birthweight ~ . - Preg.ended...37.wk 
                         - Drug.Add - Use.Alc - Use.Tob, 
                         data = opt_logistic, family = binomial, 
                         na.action = na.exclude)

model.birthweight.reduced.summary <- summary(model.birthweight.reduced)
model.birthweight.reduced.summary$coefficients %>% 
  round(digits = 3) %>% kable()

```


##### Likelihood Ratio Test
```{r, echo=FALSE}

anova(model.birthweight.reduced,model.birthweight
      , test = "LRT")

```
- $H_{0}$: There is no statistical difference between the reduced and full model.
- $H_{A}$: There is statistical difference between the reduced and full model. 
- P-value: 0.341
- We fail to reject the null hypothesis at significance level 0.05.
- The full model does not perform better and we decided to go with the simpler model that does not have alcohol, drug and tobacco use. 

##### Final Model
```{r, echo=FALSE}


# --- Goodness of Fit ---


generalhoslem::logitgof(opt_logistic$cat_birthweight,
                       model.birthweight.reduced$fitted.values)

```
- We are using a Hosmer-Lemeshow for Goodness-of-Fit
- $H_{0}$: There is no significant difference between the observed and expected values of the binary response variable across different groups or categories defined by the predictor variables.
- $H_{A}$: There is a significant difference between the observed and expected values of the binary response variable across different groups or categories defined by the predictor variables.
- We fail to reject the null hypothesis at a significance level of 0.05. This suggests that the model fits well.

##### Conclusion 
- Interpretations:
  - Intercept is not interpretable because age will never be zero and qualifying teeth will almost never be zero. 
  - When an individual has hypertension (compared to its absence), the odds of having a birthweight above 2500g are approximately 0.2746 times the odds of having a birthweight below 2500g, with other variables held constant.
  - Another way to put this is it  that the odds of having a birthweight above 2500g (compared to below 2500g) are lower for individuals with hypertension compared to those without hypertension. 